---
title: Nonparametric Differential Abundance Testing for Microbiome Counts Data
shorttitle: "DACOMP"
author:
- name: Barak Brill
  affiliation: Department of Statistics and OR, Tel Aviv University
  email: barakbri@mail.tau.ac.il
bibliography: dacomp_bib.bib
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
package: DACOMP
vignette: |
    %\VignetteIndexEntry{Nonparametric differential abundance testing for microbiome counts data}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

foo

# Model 

$$ \vec{X}_{i} | \vec{P}_{i}, N_i^X \sim multinom\left( N^{X}_i , \vec{P}_{i} \right) ,\quad\quad \vec{P}_{i}\sim\mathcal{P}\quad ,0\leq P_{i,j}, \sum_{j=1}^{m}P_{i,j} = 1 , $$


$$ \vec{Y}_{l} | \vec{Q}_{l}, N_l^Y \sim multinom\left( N^{Y}_l , \vec{Q}_{l} \right) ,\quad\quad \vec{Q}_{l}\sim\mathcal{Q}\quad ,0\leq Q_{l,j}, \sum_{j=1}^{m}Q_{l,j} = 1 .$$

$$\frac{\left(P_{v_1},P_{v_2},...,P_{v_{k}}\right)}{\sum_{k'=1}^{k} P_{v_{k'}}}\stackrel{d}{=}\frac{\left(Q_{v_1},Q_{v_2},...,Q_{v_{k}}\right)}{\sum_{k'=1}^{k} Q_{v_{k'}}} $$

$$ H_0^{(j)}:\quad \frac{\left(P_j,P_{b_1},P_{b_2},...,P_{b_{r}}\right)}{P_j + \sum_{k=1}^{r} P_{b_k}}\stackrel{d}{=}\frac{\left(Q_j,Q_{b_1},Q_{b_2},...,Q_{b_{r}}\right)}{Q_j + \sum_{k=1}^{r} Q_{b_k}}.$$


$$\tilde{H}_0^{(j)}: \tilde{X}_{i,j} \stackrel{d}{=} \tilde{Y}_{l,j} $$

## Choosing reference taxa

$$ SD_{j,k} = \mathop{\large{\mathrm{sd}}}^{n_X+n_Y}_{i=1}\left(log_{10}\left(\frac{Z_{i,j}+1}{Z_{i,k}+1}\right)\right), $$

foo

$$ S_j = \mathop{\large{\mathrm{median}}}^{m}_{k=1, k\ne j}\left(SD_{j,k}\right).$$

$$ Select\, B: \quad B = \{j|S_j \leq S_{crit}\}. $$


# Workflow
```{r,eval = F}
library(dacomp)

set.seed(1)

data = dacomp.generate_example_dataset(m1 = 100,
        n_X = 50,
        n_Y = 50,
        signal_strength_as_change_in_microbial_load = 0.1)
```

## Selecting reference taxa
```{r, eval = F}
#select references: (may take a minute)
result.selected.references = dacomp.select_references(X = data$counts,
                                                    median_SD_threshold = 0.6, #APPLICATION SPECIFIC
                                                    verbose = T)

length(result.selected.references$selected_references)

#plot the reference selection scores (can also be used to better set the median SD threshold)
dacomp.plot_reference_scores(result.selected.references)
```

## Testing for differential abundance
```{r, eval = F}
#multiplicity correction levels for the BH and DS-FDR methods
q_BH = q_DSFDR = 0.1

#Perform testing:
result.test = dacomp.test(X = data$counts,
                     y = data$group_labels,
                     ind_reference_taxa = result.selected.references,
                     verbose = T,q = q_DSFDR)

rejected_BH = which(p.adjust(result.test$p.values.test,method = 'BH')<=q_BH)
rejected_DSFDR = result.test$dsfdr_rejected
```

## Using different non parametric tests

$\verb|DACOMP.TEST.NAME.WILCOXON|$

$\verb|DACOMP.TEST.NAME.LOG_FOLD_DIFFERENCE_IN_MEANS|$

$\verb|DACOMP.TEST.NAME.TWO_PART_WILCOXON|$


```{r, eval = F}
result.test = dacomp.test(X = data$counts,
                     y = data$group_labels,
                     ind_reference_taxa = result.selected.references,
                     test = DACOMP.TEST.NAME.LOG_FOLD_DIFFERENCE_IN_MEANS,
                     verbose = T,q = q_DSFDR)

result.test = dacomp.test(X = data$counts,
                     y = data$group_labels,
                     ind_reference_taxa = result.selected.references,
                     test = DACOMP.TEST.NAME.DIFFERENCE_IN_MEANS,
                     verbose = T,q = q_DSFDR)

result.test = dacomp.test(X = data$counts,
                     y = data$group_labels,
                     ind_reference_taxa = result.selected.references,
                     test = DACOMP.TEST.NAME.TWO_PART_WILCOXON,
                     verbose = T,q = q_DSFDR)
```

# Testing with $K$ sample groups
$\verb|DACOMP.TEST.NAME.KRUSKAL_WALLIS|$

# Testing with paired study design
$\verb|DACOMP.TEST.NAME.WILCOXON_SIGNED_RANK_TEST|$

```{r, eval = F}
set.seed(1)
data = dacomp.generate_example_dataset_paired(30)

#image(t(log10(data$counts+1)),xlab = 'taxa',ylab = 'subject')
#abline(h = 0.5,col = 'black')


data$counts

result.selected.references = dacomp.select_references(X = data$counts,
                                                      median_SD_threshold = 0.6, #APPLICATION SPECIFIC
                                                      verbose = T)


length(result.selected.references$selected_references)

#plot the reference selection scores (can also be used to better set the median SD threshold)
dacomp.plot_reference_scores(result.selected.references)


#multiplicity correction levels for the BH and DS-FDR methods
q_BH = q_DSFDR = 0.1

#Perform testing:
result.test = dacomp.test(X = data$counts,
                          y = NULL,
                          ind_reference_taxa = result.selected.references, test = DACOMP.TEST.NAME.WILCOXON_SIGNED_RANK_TEST,
                          verbose = T,q = q_DSFDR)

rejected_BH = which(p.adjust(result.test$p.values.test,method = 'BH')<=q_BH)
rejected_DSFDR = result.test$dsfdr_rejected

sum(rejected_DSFDR %in% data$select_diff_abundant) 
length(rejected_DSFDR)

sum(rejected_BH %in% data$select_diff_abundant) 
length(rejected_BH)
```


# Testing with a continous covariate
$\verb|DACOMP.TEST.NAME.SPEARMAN|$

```{r,eval = F}
set.seed(1)
data = dacomp.generate_example_dataset_continuous(n = 100,m1 = 30,
signal_strength_as_change_in_microbial_load = 0.1)


result.selected.references = dacomp.select_references(X = data$counts,
                                                      median_SD_threshold = 0.6, #APPLICATION SPECIFIC
                                                      verbose = T)
#number of selected references
length(result.selected.references$selected_references)
#plot the reference selection scores (can also be used to better set the median SD threshold)
dacomp.plot_reference_scores(result.selected.references)

#multiplicity correction levels for the BH and DS-FDR methods
q_BH = q_DSFDR = 0.1

#Perform testing:
result.test = dacomp.test(X = data$counts,
                          y = data$covariate,test = DACOMP.TEST.NAME.SPEARMAN,
                          ind_reference_taxa = result.selected.references,
                          verbose = T,q = q_DSFDR)

rejected_BH = which(p.adjust(result.test$p.values.test,method = 'BH')<=q_BH)
rejected_DSFDR = result.test$dsfdr_rejected
 
```


# Development

## Other reference selection methods
  
## User defined tests

```{r,eval = F}
set.seed(1)

data = dacomp.generate_example_dataset_multivariate_example(n = 100,m1 = 30,signal_strength_as_change_in_microbial_load = 0.1)

#two covariates for each subject
head(data$covariate)
#            u1         u2
#[1,] 0.4820801 0.57487220
#[2,] 0.5995658 0.07706438
#[3,] 0.4935413 0.03554058
#[4,] 0.1862176 0.64279549
#[5,] 0.8273733 0.92861520
#[6,] 0.6684667 0.59809242

result.selected.references = dacomp.select_references(X = data$counts,
                                                      median_SD_threshold = 0.5,
                                                      verbose = T)



#multiplicity correction levels for the BH and DS-FDR methods
q_BH = q_DSFDR = 0.1


nr_perm_to_perform = 1000

library(energy)
custom_test_function = function(X){
  res = dcov.test(X, data$covariate, R=nr_perm_to_perform)
  return(
          c(
            res$statistic ,
            res$replicates
            )
    )
}

#Perform testing:
result.test = dacomp.test(X = data$counts,
                          y = NULL,
                          test = DACOMP.TEST.NAME.USER_DEFINED,
                          ind_reference_taxa = result.selected.references,
                          verbose = T,q = q_DSFDR,
                          nr_perm = nr_perm_to_perform, # should be exactly the same as the 
                          user_defined_test_function = custom_test_function)

rejected_BH = which(p.adjust(result.test$p.values.test,method = 'BH')<=q_BH)
rejected_DSFDR = result.test$dsfdr_rejected


```

  